---
title: "Q96 analysis 2.0"
output: html_notebook
---

# Packages and directories
```{r eval=FALSE}
library(dplyr)
library(tidyr)
library(lattice)
library(janitor)
library(vroom)
library(ggplot2)
library(viridis)
library(ggpubr)
library(tidyverse)

inputs <- paste0(getwd(), "/inputs/")
outputs <- paste0(getwd(), "/outputs/")
```

# Import and clean

```{r}
soars_b_pheno <- vroom(paste0(inputs, "sb_pheno_data_20210123.txt")) %>% clean_names()
oxtr_meth <- vroom(paste0(inputs,"20210202_Q96_QC_filtered_samples.txt")) %>% 
  clean_names() %>%
  dplyr::select(sample_id, mean_901, mean_924, mean_934, batch) %>%
  rowwise() %>%
  mutate(avg_meth_total = mean(c(mean_901, mean_924, mean_934), na.rm = TRUE))

# Combine OXTR methylation data with the paricipant demographics / sample / plasmsa OT table
analysis_df_base <- left_join(oxtr_meth, soars_b_pheno, by = c("sample_id" = "acquisition_id")) %>%
  filter(!is.na(subject)) %>%
  filter(sample_id != "A2000527")

analysis_df_base$funstrat <- as.factor(analysis_df_base$funstrat)

# Need to convert timepoints into a standardized format like Steph used 

table(analysis_df_base$study_timepoint)

# Doing this using recode
analysis_df_base <- analysis_df_base %>%
  mutate(study_timepoint = recode(study_timepoint,
                                  "Baseline" = "baseline",
                                  "-1" = "baseline",
                                  "0" = "baseline",
                                  "24" = "week_24",
                                  "36" = "week_36",
                                  "8" = "week_8",
                                  "8/ET" = "week_8",
                                  "Baseline_Week 0" = "baseline",
                                  "SCR" = "screen",
                                  "Screen" = "screen",
                                  "Scrn" = "screen",
                                  "scr" = "screen",
                                  "W24" = "week_24",
                                  "W36" = "week_36",
                                  "W8" = "week_8",
                                  "32" = "week_36",
                                  "40" = "week_36",
                                  "WK 36" = "week_36",
                                  "Week 24" = "week_24",
                                  "Week 36" = "week_36",
                                  "Week 8" = "week_8",
                                  "Wk 24" = "week_24",
                                  "Wk 36" = "week_36",
                                  "Wk 40" = "week_40",
                                  "Wk 48" = "week_48",
                                  "Wk 8" = "week_8"))
table(analysis_df_base$study_timepoint)


# Designate factors where appropriate
colnames(analysis_df_base) 

analysis_df_base <- analysis_df_base %>%
  mutate_at(vars(subject, randoid, agestrat, funstrat, male, hispanic, race_d, 
                 site_r, tx, race_d3, assay_batch, assay_batch_plate, oxt_batch_num), 
            list(factor)) %>%
  rename("ot_batch" = "oxt_batch_num")

# If a sample has both a baseline and a screen sample, select baseline
df_baseline <- filter(analysis_df_base, study_timepoint == "baseline" | study_timepoint == "screen")
df_baseline_dups <- df_baseline %>% 
  group_by(subject) %>% 
  filter(n() > 1)

df_baseline <- df_baseline %>%
  filter(!sample_id %in% df_baseline_dups$sample_id)

df_baseline <- bind_rows(df_baseline, filter(df_baseline_dups, study_timepoint == "baseline"))

# For use in the longitudinal plots
df_long <- analysis_df_base %>%
  pivot_longer(cols = starts_with("mean"), names_to = "cpg_site")


```

# Subsetting to baseline and looking at associations with plasma oxytocin and other demographics data
```{r}
#### Average methylation values across all baseline samples and batches

my_theme <- theme_bw() +
  theme(axis.text.x = element_text(hjust = 0.5, size = 14, angle = 0, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 14, face = "bold"),
        strip.text.x = element_text(size = 14, face = "bold"),
        strip.text.y = element_text(size = 14, face = "bold"),
        strip.background = element_blank())



batch_meth <- ggplot(data = oxtr_meth, aes(x = batch, y = avg_meth_total)) +
  geom_boxplot(width = 0.5, size = 0.8) + 
  geom_point() +
  my_theme +
  ggtitle("Average Methylation Across -901, -924, -934") +
  ylim(25,75) + ylab("% Methylation") + xlab("Q96 batch")
batch_meth

ggsave(filename = paste0(outputs, "pyrosequencing_batch_average_meth_20210223.png"),
       plot = batch_meth, dpi = 200, width = 10, height = 8, units = "in")
```


# Spaghettii plot of avg meth over time for subject
```{r}

analysis_df_base$study_timepoint <- factor(analysis_df_base$study_timepoint, 
                                           levels = c("screen", "baseline", "week_8",
                                                      "week_24", "week_36", "week_40", 
                                                      "week_48"))
analysis_df_base$funstrat <- as.factor(analysis_df_base$funstrat)
spaghetti_avg <- analysis_df_base %>%
  filter(study_timepoint != "week_40" & study_timepoint != "week_48") %>%
  group_by(funstrat, study_timepoint) %>%
  summarise(avg_meth_total = mean(avg_meth_total))

spaghetti_meth <- ggplot(data = filter(analysis_df_base, study_timepoint != "week_40" &
                                         study_timepoint != "week_48"), 
                         aes(x = study_timepoint, y = avg_meth_total, group = subject, color = funstrat)) +
  my_theme +
  theme(legend.position = "bottom", 
        legend.title = element_blank(), 
        title = ggtext::element_markdown(),
        legend.text=element_text(size = 18)) +
  ggtitle("Change in Average OXTR Methylation Over Time") +
  ylab("% Methylation") + 
  xlab("") +
  ylim(25,70) +
  scale_x_discrete(labels=c("screen", "baseline", "week-8", "week-24", "week-36")) +
  geom_line(size = 1, alpha = 0.1) +
  geom_line(data = spaghetti_avg, 
            aes(x = study_timepoint, y = avg_meth_total, group = funstrat, color = funstrat),
            size = 3, alpha = 0.8) + 
  scale_color_viridis(discrete = TRUE, option = "D", begin = .1, end = .8, labels = c("Minimally Verbal", "Verbal"))
  #scale_fill_viridis(discrete = TRUE, option = "A", begin = 0.3, end = 0.8, labels = c("Minimally Verbal", "Verbal")) 


ggsave(filename = paste0(outputs, "methylation_over_time_avg_202102224.png"),
       plot = spaghetti_meth, dpi = 200, width = 12, height = 6, units = "in")
```

# Linear Models
```{r}
ggplotRegression <- function (fit, xlab, ylab, title) {

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  my_theme + 
  geom_point() +
  geom_abline(slope = coef(fit)[[2]], intercept = coef(fit)[[1]], col = "blue") +
  geom_smooth(method = "lm", col = "red") +
  ylim(c(0, 100)) +
  labs(title = title,
       subtitle = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                        "Intercept =",signif(fit$coef[[1]],5 ),
                        " Slope =",signif(fit$coef[[2]], 5),
                        " P =",signif(summary(fit)$coef[2,4], 5))) +
  xlab(xlab) +
  ylab(ylab)
}

ggplotRegression_box <- function (fit, xlab, ylab, title) {

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  my_theme + 
  geom_boxplot(alpha = 0.5) +
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(title = title,
       subtitle = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                        "Intercept =",signif(fit$coef[[1]],5 ),
                        " Slope =",signif(fit$coef[[2]], 5),
                        " P =",signif(summary(fit)$coef[2,4], 5))) +
  xlab(xlab) +
  ylab(ylab)
}
```

# Inspecting the variables
```{r}
library(leaps)
library(relaimpo)

# Leaps
leaps <- regsubsets(avg_meth_total ~ log_ot_num + batch + male + study_site + 
            funstrat + ageyrs + pc1 + pc2 + pc3 + ot_batch, 
            data = df_baseline, nbest=10)

# view results
summary(leaps)

# plot a table of models showing variables in each model.
# models are ordered by the selection statistic.

png(filename = paste0(outputs, "OXTR_avg_meth_factor_analysis_20210303.png"), 
    width = 10, height = 6, units = "in", res = 100)
plot(leaps,scale="r2", )
dev.off()

# plot statistic by subset size

subsets(leaps, statistic="rsq")

# Relative imporatance
library(relaimpo)
calc.relimp(fit, type=c("lmg", "last", "first"),
   rela=TRUE)

# Bootstrap Measures of Relative Importance (1000 samples)
boot <- boot.relimp(fit, b = 1000, type = c("lmg"), rank = TRUE,
  diff = TRUE, rela = TRUE)
booteval.relimp(boot) # print result
plot(booteval.relimp(boot,sort=TRUE)) # plot result 

```

#### Overall methylation
```{r}
fit <- lm(avg_meth_total ~ log_ot_num + batch + male + study_site + 
            funstrat + ageyrs + pc1 + pc2 + pc3 + ot_batch, 
          data = df_baseline)
summary(fit)

lm_overall_meth <- ggplotRegression(fit = fit, title = "Average OXTR methylation & Plasma Oxytocin",
                                    xlab = "LN Plasma Oxytocin",
                                    ylab = "% Methylation")

```
# Individual CpG Meth & pOT

#### -901 methylation
```{r}

fit <- lm(mean_901 ~ log_ot_num + batch + male + study_site + 
            funstrat + ageyrs + pc1 + pc2 + pc3 + ot_batch, data = df_baseline)
summary(fit)
lm_901_meth <- ggplotRegression(fit = fit, title = "Average OXTR -901 methylation & Plasma Oxytocin",
                                    xlab = "LN Plasma Oxytocin",
                                    ylab = "% Methylation")
```

#### -924 methylation
```{r}

fit <- lm(mean_924 ~ log_ot_num + batch + male + study_site + 
            funstrat + ageyrs + pc1 + pc2 + pc3 + ot_batch, 
          data = df_baseline)
summary(fit)
lm_924_meth <- ggplotRegression(fit = fit, title = "Average OXTR -924 methylation & Plasma Oxytocin",
                                    xlab = "LN Plasma Oxytocin",
                                    ylab = "% Methylation")

```

#### -934 methylation
```{r}

fit <- lm(mean_934 ~ log_ot_num + funstrat + batch + male + ageyrs + pc1 + pc2 + pc3 + ot_batch, data = df_baseline)

lm_934_meth <- ggplotRegression(fit = fit, title = "Average OXTR -934 methylation & Plasma Oxytocin",
                                    xlab = "LN Plasma Oxytocin",
                                    ylab = "% Methylation")
```

#### LM Multiplot
```{r}
library(cowplot)

lm_grid_meth_ot <- plot_grid(lm_overall_meth + theme(axis.title.x = element_blank()),
                     lm_901_meth + theme(axis.title.x = element_blank(), 
                                         axis.title.y = element_blank()),
                     lm_924_meth,
                     lm_934_meth + theme(axis.title.y = element_blank()),
                     ncol = 2,
                     nrow = 2, align = "hv", labels = c("A","B","C","D"))

ggsave(filename = paste0(outputs, "meth_linear_models_plasma_ot_202102224.png"),
       plot = lm_grid_meth_ot, dpi = 100, width = 12, height = 12, units = "in")
```


# Is there an association between methylation in OXTR and functional strata?


#### Overall methylation
```{r}
fit <- lm(avg_meth_total ~ funstrat + log_ot_num + batch + male + ageyrs + pc1 + pc2 + pc3 + oxt_batch, 
          data = df_baseline)
lm_overall_meth_fs <- ggplotRegression_box(fit = fit, title = "Average OXTR methylation & Functional Strata",
                                    xlab = "Functional Strata",
                                    ylab = "% Methylation")
lm_overall_meth_fs <- lm_overall_meth_fs + scale_x_discrete(labels=c("Non-verbal", "Verbal"))
```
# Individual CpG Meth & pOT

#### -901 methylation
```{r}

fit <- lm(mean_901 ~ funstrat + log_ot_num + batch + male + ageyrs + pc1 + pc2 + pc3 + oxt_batch, 
          data = df_baseline)

lm_901_meth_fs <- ggplotRegression_box(fit = fit, title = "Average OXTR -901 methylation & Functional Strata",
                                    xlab = "Functional Strata",
                                    ylab = "% Methylation")

lm_901_meth_fs <- lm_901_meth_fs + scale_x_discrete(labels=c("Non-verbal", "Verbal"))
```

#### -924 methylation
```{r}

fit <- lm(mean_924 ~ funstrat + log_ot_num + batch + male + ageyrs + pc1 + pc2 + pc3 + oxt_batch, data = df_baseline)

lm_924_meth_fs <- ggplotRegression_box(fit = fit, title = "Average OXTR -924 methylation & Functional Strata",
                                    xlab = "Functional Strata",
                                    ylab = "% Methylation")
lm_924_meth_fs <- lm_924_meth_fs + scale_x_discrete(labels=c("Non-verbal", "Verbal"))
```

#### -934 methylation
```{r}

fit <- lm(mean_934 ~ funstrat + log_ot_num + batch + male + ageyrs + pc1 + pc2 + pc3 + oxt_batch, data = df_baseline)

lm_934_meth_fs <- ggplotRegression_box(fit = fit, title = "Average OXTR -934 methylation & Functional Strata",
                                    xlab = "Functional Strata",
                                    ylab = "% Methylation")
lm_934_meth_fs <- lm_934_meth_fs + scale_x_discrete(labels=c("Non-verbal", "Verbal"))
```

#### LM_funstrat Multiplot
```{r}
lm_grid_meth_fs <- plot_grid(lm_overall_meth_fs + theme(axis.title.x = element_blank()),
                     lm_901_meth_fs + theme(axis.title.x = element_blank(), 
                                         axis.title.y = element_blank()),
                     lm_924_meth_fs,
                     lm_934_meth_fs + theme(axis.title.y = element_blank()),
                     ncol = 2,
                     nrow = 2, align = "hv", labels = c("A","B","C","D"))

ggsave(filename = paste0(outputs, "meth_linear_models_functional_strata_202102224.png"),
       plot = lm_grid_meth_fs, dpi = 200, width = 12, height = 12, units = "in")

```

```{r}
plot <- ggplot(df_baseline, aes(x = funstrat, 
                                       y = log_ot_num,
                                       na.rm = TRUE)) + 
  geom_boxplot(lwd = 1.5, outlier.shape = NA, aes(color = funstrat)) +
  geom_point(shape = 21, size = 4, stroke = 2, aes(color = funstrat)) +
  scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8, option = "D") +
  # geom_signif(comparisons = list(c("high_functioning", "low_functioning")),
  #             map_signif_level = TRUE) +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, size = 30), legend.title = element_blank(),
        axis.text = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(hjust = 0.5, size = 24),
        legend.text = element_text(size = 15, face = "bold"),
        axis.line = element_line(colour = 'black', size = 1.2),
        axis.ticks = element_line(colour = "black", size = 1.2)) +
  labs(y = "Plasma OT (log)", x = "")
 

plot

ggsave("ot_funstrat.png", plot = plot, path = outputs, device = "png", 
        dpi = 300, width = 10, height = 8, units = "in")
```

# Interaction terms
```{r}

plot(avg_924 ~ log_ot_num, combined_tables_filtered, pch=19)
fit <- lm(avg_924 ~ log_ot_num:male:funstrat + ageyrs + batch + pc1 + pc2 + pc3 + oxt_batch, data = combined_tables_filtered)
abline(fit)
summary(fit)
plot(fit)

```

# Plotting the methylation controls

```{r}
control_subset <- filter(oxtr_meth, sample_id == "0%_MC" | sample_id == "50%_MC" | sample_id == "100%_MC")

plot <- ggplot(control_subset, aes(x = sample_id, y = avg_meth_total, na.rm = TRUE)) + 
  geom_boxplot(lwd = 1.5, outlier.shape = NA, aes(color = sample_id)) +
  geom_point(shape = 21, size = 4, stroke = 2, aes(color = sample_id)) +
  scale_color_viridis(discrete = TRUE, begin = 0.2, end = 0.8, option = "D") +
  # geom_signif(comparisons = list(c("high_functioning", "low_functioning")),
  #             map_signif_level = TRUE) +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, size = 30), legend.title = element_blank(),
        axis.text = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(hjust = 0.5, size = 24),
        legend.text = element_text(size = 15, face = "bold"),
        axis.line = element_line(colour = 'black', size = 1.2),
        axis.ticks = element_line(colour = "black", size = 1.2)) +
  labs(y = "Plasma OT (log)", x = "")
 

plot

ggsave("ot_funstrat.png", plot = plot, path = outputs, device = "png", 
        dpi = 300, width = 10, height = 8, units = "in")
```



