---
title: "Q96 Rerun Sample Management"
output: html_notebook
---

```{r echo=FALSE}
library(dplyr)
library(vroom)
library(stringr)
library(janitor)
```

```{r}
inputs <- paste0(getwd(), "/inputs/")
outputs <- paste0(getwd(), "/outputs/")
```

# Read-in & Process Files
```{r}

rerun_find_subs <- rerun_qc_results %>%
  select(sample_id, genolist_plate) %>%
  left_join(select(all_allocations,
                   sample, tube_barcode),
            by = c("sample_id" = "sample"))
  
# Any in the list that are not in the box? 
rerun_find_subs <- left_join(rerun_find_subs, 
                          rerun_plates, by = c("tube_barcode" = "code")) %>%
  arrange(plate, column, row)

# A2005544

write.table(rerun_find_subs,
            file = paste0(outputs, "20210202_rerun_sample_list.txt"),
            quote = FALSE, row.names = FALSE, col.names = TRUE, sep = "\t")

```

# Post volume-check quant
I visually inspected the samples that need to be rerun and were not in plates J or K, consolidated them, and rescanned them. Linking the scan file to nanodrop results. 

```{r}
inspected_tubes <- read.csv(paste0(inputs, "20210204_Q96_rerun_tubes.txt"),
                            sep = "\t") %>% clean_names() %>%
  left_join(select(all_allocations, sample, tube_barcode),
            by = c("code" = "tube_barcode"))

write.table(select(inspected_tubes, sample), file = paste0(outputs, "20210204_nanodrop_sample_names.txt"), sep = "\t", row.names = FALSE,
            col.names = TRUE, quote = FALSE)


# Importing Q96 nano results

inspected_tubes <- inspected_tubes %>%
  left_join(select(nano_values, sample_id, conc, x260_280, x260_230),
            by = c("sample" = "sample_id")) %>%
  filter(!is.na(sample))

reallocations <- filter(inspected_tubes, conc < 10)
```

# Alignment of pass/failed samples with genolist

```{r}
passed_samples <- vroom(paste0(inputs, "20210202_Q96_QC_filtered_samples.txt")) %>%
  filter(!sample_id %in% c("0%_MC", "50%_MC", "100%_MC", "NTC", "blank")) %>%
  mutate(status = "passed_QC")

failed_samples <- vroom(paste0(inputs, "20210202_Q96_failed_QC_samples.txt")) %>%
  filter(!sample_id %in% c("0%_MC", "50%_MC", "100%_MC", "NTC", "blank")) %>%
  mutate(status = "failed_QC")

all_run_samples <- bind_rows(passed_samples, failed_samples)
all_allocations <- vroom(paste0(inputs, "allDNAalloc.26Jan2021.txt")) %>% clean_names()

genolist <- vroom(paste0(inputs, "genolist.txt")) %>% clean_names() %>%
  rename("genolist_plate" = "plate_96")

genolist_sub_QC_blank <- genolist %>%
  filter(sample != "blank" & !str_detect(sample, ".*QC.*"))

genolist_results <- left_join(genolist_sub_QC_blank,
                              all_run_samples,
                              by = c("sample" = "sample_id"))

# These are the samples I need to convert
rerun_samples <- filter(genolist_results, status == "failed_QC")
rerun_BSC <- filter(rerun_samples, !genolist_plate %in% c("J", "K")) %>%
  select(sample)

# Find where they are stored

nrow(filter(nano_values, sample_id %in% rerun_BSC$sample))
# All 72 samples in the nanodropped rerun plate are in the post-QC rerun list
# (as expected)

rerun_BSC_find <- filter(rerun_BSC, !sample %in% nano_values$sample_id)
# 19 samples left to track down 

# load in all of the stored samples and the genotype plates
stored_samples <- vroom(paste0(inputs, "20210205_soarsb_stored_samples.txt"))
genotype_boxes <- vroom(paste0(inputs, "final_gt_plate_scans_20210129.txt")) %>%
  rename_with(~ gsub("gt_", "", .x), starts_with("gt_")) %>%
  select(colnames(stored_samples))
rerun_box_2 <- read.csv(paste0(inputs, "20210204_Q96_rerun_tubes_2.txt"),
                        sep = "\t") %>% clean_names() %>%
  mutate(well = paste0(row, column)) %>%
  select(code, well) %>%
  left_join(select(all_allocations, sample, tube_barcode), by = c("code" = "tube_barcode"))

stored_samples <- rbind(stored_samples, genotype_boxes)


# Find the remaining samples, none in storage boxes so the rest must be in rerun plate 2
rerun_BSC_find <- rerun_BSC_find %>%
  left_join(rerun_box_2, by = "sample") %>%
  mutate(box = "rerun_2")

write.table(rerun_BSC_find, file = paste0(outputs, "20210204_rerun19_nano.txt"),
            sep = "\t", col.names = TRUE, row.names = FALSE, quote = FALSE)
```


```{r}
# nanodrop
nano_values <- read.csv(paste0(inputs, "20210204_q96_rerun_nano.txt"),
                        sep = "\t") %>% clean_names() %>%
  select(sample_id, well, conc) %>%
  left_join(rerun_BSC, keep = TRUE, by = c("sample_id" = "sample"))

nano_values_2 <- read.csv(paste0(inputs, "20210204_q96_rerun_nano_2.txt"),
                        sep = "\t") %>% clean_names() %>%
  select(sample_id, well, conc) %>%
  left_join(rerun_BSC, keep = TRUE, by = c("sample_id" = "sample"))

# There were some duplicate tubes in the nano file because I wasn't sure which tube would be viable. Identify those, remove the low conc alternative

nano_values %>% group_by(sample_id) %>% filter(n() > 1)
# H2-A2001830 kept
# F3-A2001182 kept
# F5-A2001160 kept
# F9-A2003289 kept

nano_values_updated <- filter(nano_values, !well %in% c("B3", "E3", "G5", "G9"))
nano_values_updated <- bind_rows(nano_values_updated, nano_values_2) %>%
  select(sample, well, conc)

failed_samples_nano <- failed_samples %>%
  left_join(nano_values_updated, by = c("sample_id" = "sample"))

filter(rerun_samples, !sample %in% nano_values_updated$sample)

# All bases covered. Of the samples that failed QC in the gennolist results file, all are represented on the current rerun plate and the bisulfite converted DNA from batches J and K. 

# Now I just need to request allocations for any sample with insufficient quantity
write.table(filter(nano_values_updated, conc < 20), 
            file = paste0(outputs,"20210205_reallocation_check.txt"),
            sep = "\t", row.names = FALSE, col.names = TRUE,
            quote = FALSE)
filter(nano_values_updated, conc < 20)
```

Visually inspecting those flagged samples

sample	well	conc  keep
A2000648	A2	2.84  no
A2002485	C2	15.47 yes
A2000653	E2	-8.035  no
A2003806	C3	7.731 yes
A2003842	H3	13.87 yes
A2000557	F4	10.76 yes
A2000645	G4	12.15 yes
A2001160	F5	4.351 no
A2000570	B6	12.12 yes
A2003035	E7	-0.8811 yes (super high vol)
A2003048	F7	-0.6495 yes (super high vol)
A2001098	A8	15.95 yes
A2001106	E8	16.8  yes

Emailing steph to request reallocations where necessary
